Data Science Applied to Ag - Final Project - ML

### 2025-04-25 code/Draft_script.qmd
d

@@

library(stringr)

@@

summary(soil_clean)
```

@@

@@

meta_clean <- meta_df %>%
  rename(
    lon = longitude,
    lat = latitude
  ) %>%
  mutate(
    # 1) remove all lowercase letters
    site = str_remove_all(site, "[a-z]"),
    # 2) keep only text before any dash
    site = str_replace(site, "-.*$", ""),
    # 3) keep only text before any underscore
    site = str_replace(site, "_.*$", "")
  ) %>%
  distinct(year, site, .keep_all = TRUE) %>%
  ungroup()

@@

```{r}
library(stringr)

meta_clean <- meta_clean %>%
  mutate(
    # 1) lowercase & trim
    prev_crop_raw = str_to_lower(str_squish(previous_crop)),
    # 2) recode into your desired levels
    prev_crop = case_when(
      prev_crop_raw == "cotton"                                   ~ "cotton",
      prev_crop_raw == "peanut"                                   ~ "peanut",
      prev_crop_raw %in% c("soybean", "soybeans")                 ~ "soybean",
      prev_crop_raw == "sorghum"                                  ~ "sorghum",
      prev_crop_raw %in% c("wheat", "winter wheat", "2019/20 wheat") ~ "wheat",
      TRUE                                                         ~ "others"
    ) %>%
    # 3) make it a factor with the exact ordering you want
    factor(levels = c("cotton", "peanut", "soybean", "sorghum", "wheat", "others"))
  ) %>%
  select(-previous_crop, -prev_crop_raw) %>%
  clean_names()

@@

@@

@@

merged_clean  <- trait_avg_df %>%
  left_join(soil_clean, by = c("year", "site")) %>%
  left_join(meta_clean, by = c("year", "site"))

summary(merged_clean)
```

# Exporting

The following code chunk will export the wrangled data into a .csv.

@@

@@

states <- us_states() %>% 
  filter( !(state_abbr %in% c("PR", "AK", "HI")) ) #to remove "PR" (Puerto Rico), "AK" (Alaska), and "HI" (Hawaii) from the rows of "states" object i.e., to keep all states that are NOT "PR", "AK", and "HI" #we use "%in%" to filter (i.e., work on) more than one state/entry #if we wanted to filter just 1 state, we would use == sign e.g., !(state_abbr == c("PR"))

@@

@@

@@

# Extract unique combinations of year, site, and their coordinates
site_year_df <- merged_daymet %>%
  select(year, site, lon, lat) %>%  # include longitude and latitude
  distinct() %>%                     # one row per unique combination
  arrange(year, site)

# Inspect the result
site_year_df

@@

weather_daymet_unnest <- weather_daymet_all %>%
  select(year, site, weather) %>% 
  unnest(weather) %>% 
  pivot_wider(names_from = measurement, 
              values_from = value) %>% 
  janitor::clean_names() #

>>>>

trait_clean

@@

@@

soil_clean
```

@@

# Data wrangling on "meta_df"

@@

meta_df2 <- meta_df %>%
  rename(
    lon = longitude,
    lat = latitude
  ) %>%
  mutate(
    # 1) remove all lowercase letters
    site = str_remove_all(site, "[a-z]"),
    # 2) keep only text before any dash
    site = str_replace(site, "-.*$", ""),
    # 3) keep only text before any underscore
    site = str_replace(site, "_.*$", "")
  ) %>%
  distinct(year, site, .keep_all = TRUE) %>%
  ungroup()

meta_df2

@@

meta_clean <- meta_df2 %>%
  mutate(
    # 1) lowercase & trim
    prev_crop_raw = str_to_lower(str_squish(previous_crop)),
    # 2) recode into your desired levels
    prev_crop = case_when(
      prev_crop_raw == "cotton"                                   ~ "cotton",
      prev_crop_raw == "peanut"                                   ~ "peanut",
      prev_crop_raw %in% c("soybean", "soybeans")                 ~ "soybean",
      prev_crop_raw == "sorghum"                                  ~ "sorghum",
      prev_crop_raw %in% c("wheat", "winter wheat", "2019/20 wheat") ~ "wheat",
      TRUE                                                         ~ "others"
    ) %>%
    # 3) make it a factor with the exact ordering you want
    factor(levels = c("cotton", "peanut", "soybean", "sorghum", "wheat", "others"))
  ) %>%
  select(-previous_crop, -prev_crop_raw) %>%
  clean_names()

@@

# EDA for cleaned data

```{r}
summary(trait_clean)
```
```{r}
summary(soil_clean)
```

@@

# Merging all 3 cleaned data frames

@@

merged_clean  <- trait_clean %>%
  left_join(soil_clean, by = c("year", "site")) %>%
  left_join(meta_clean, by = c("year", "site"))

merged_clean
```

```{r}
summary(merged_clean)
```

```{r exporting merged data set}

@@

The following code chunk will create a map of the USA and plot the sites in the map based on their latitude and longitude.

@@

states <- us_states() %>% 
  filter( !(state_abbr %in% c("PR", "AK", "HI")) ) #to remove "PR" (Puerto Rico), "AK" (Alaska), and "HI" (Hawaii)

@@

The following code chunk will keep the observations for site-years having latitude and longitude withing the Daymet range of co-ordinates.

@@

```
The following code chunk will extract unique combinations of year, site, and their coordinates.

@@

site_year_df <- merged_daymet %>%
  select(year, site, lon, lat) %>%  # need to include longitude and latitude along with site-years
  distinct() %>%                    
  arrange(year, site)

site_year_df

@@

weather_daymet_unnest <- weather_daymet_all %>%
  select(year, site, weather) %>% 
  unnest(weather) %>% 
  pivot_wider(names_from = measurement, 
              values_from = value) %>% 
  janitor::clean_names()
